---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# The `h3sampling` R package

<!-- badges: start -->
<!-- badges: end -->

The `h3sampling` package implements spatially balanced sampling designs which leverage the [H3](https://h3geo.org/) discrete global grid system for indexing spatial data into a hexagonal grid.
The package currently only provides spatially balanced sampling via an adaptation of the Generalized Random Tessellation Stratified (GRTS) algorithm.
By making use of the [H3](https://h3geo.org/) global grid, the package is especially suitable for so-called master samples.
For a given H3 grid resolution and a given random seed, all H3 cells are deterministically sorted such that consecutively ordered cells are spatially balanced samples for a given area on earth.

The implementation is written in [`Rust`](https://rust-lang.org/) with the aid of `rextendr`.
Large language models Google Gemini pro and Antropic Claude Sonnet 4.6 were used.
The former for brainstorming and code writing, the latter for a final critic and improvements on the code base.
R code is kept to a bare minimum and the package has zero R dependencies.
This results in a lightweight, cross-platform, fast and memory friendly implementation.

## Installation

To install the development version of `h3sampling`:

``` r
remotes::install_github("inbo/h3sampling")
```

## Example

For an example, we extract Ashe county from the North Caroline state dataset that ships with the `sf` package.
We draw a spatially balanced sample with the aid of `h3_grts()` at resolution 7 of the [H3 resolutions](https://h3geo.org/docs/core-library/restable) system.
This corresponds to hexagonal areas of approximately 5.16 km².
The `h3o` package can be used to, among other things, obtain the hexagon centroid or vertices.


```{r example}
library(h3sampling)
nc_path <- system.file("shape/nc.shp", package = "sf")
ashe <- sf::st_read(
  nc_path,
  quiet = TRUE,
  query = "SELECT * FROM nc WHERE NAME='Ashe'"
)
ashe_wkb <- ashe |>
  sf::st_geometry() |>
  sf::st_as_binary()
ashe_wkb <- ashe_wkb[[1]]
  
h3_sample_cells <- h3_grts(
  wkb = ashe_wkb,
  n = 20,
  resolution = 7,
  containment = "centroid",
  seed = 123
)
h3_sample_points <- h3_sample_cells |>
  h3o::h3_from_strings() |>
  h3o::h3_to_points()
h3_sample_hexagons <- h3_sample_cells |>
  h3o::h3_from_strings() |> 
  h3o::h3_to_vertexes() |>
  sf::st_cast("POLYGON")
```

We can visualise the selected spatially balanced sample and the H3 cells:

```{r ashe-example}
ashe_cells <- h3o::sfc_to_cells(
  ashe$`_ogr_geometry_`,
  resolution = 7,
  containment = "centroid"
)
ashe_hexagons <- h3o::h3_to_vertexes(ashe_cells[[1]]) |>
  sf::st_cast("POLYGON")
plot(ashe$`_ogr_geometry_`)
plot(ashe_hexagons, add = TRUE)
plot(h3_sample_hexagons, col = "lightgreen", add = TRUE)
plot(h3_sample_points, add = TRUE, cex = 0.3)
```

